---
title: "tidying_data.Rmd"
author: "Sebastian W. Ertl"
date: "Monday, February 16, 2015"
output: html_document
---

Aktueller Stand: 20.2.2015

# Data Tidying

## Introduction

In the first script "gather\_data.Rmd", I read the data from the xlsx files and save the cluttered list to xlsx\_data.rds

Now I have to check the numbers and gather some more, where they weren't read properly (in gather\_data.Rmd). Additionally the numbers must be labeled as numbers.

```{r}
library(stringr)
my_data <- as.data.frame(readRDS("xlsx_data.rds"), stringsAsFactors = FALSE)
par(mar=c(5.1,4.1,4.1,2.1))
```

In summary there are `r length(my_data[[1]])` data entries.

## Barthel-Index
```{r}
my_data$barthel <- as.numeric(as.character(my_data$barthel))
summary(my_data$barthel)
```

`r sum(is.na(my_data$barthel))` NA's, checked, all also NA in original Excel-Sheets.

## Konferenzbeschluss
So früh in der Analyse, da das Datum des Konferenzbeschlusses für die Berechnung des Alters benötigt wird.
```{r}
my_data$konferenz_bak <- my_data$konferenz
my_data$konferenz <- str_replace(my_data$konferenz, "Konferenzbeschluss vom", "")
my_data$konferenz <- str_replace(my_data$konferenz, " ", "")
my_data$konferenz <- str_replace(my_data$konferenz, "11\\.09\\.\\.14",
                                 "11.9.2014")
my_data$konferenz <- as.Date(my_data$konferenz, format = "%d.%m.%Y")
my_data$konferenz[is.na(my_data$konferenz)] <-
    as.Date(as.integer(my_data$konferenz_bak[is.na(my_data$konferenz)]),
            origin = "1899-12-30")

my_data$aufnahme <- as.Date(as.integer(my_data$aufnahme), origin = "1899-12-30")
```

```{r}
my_data$beschluss <- as.character(my_data$beschluss)

tidy_beschluss <- function(a_string) {
    ret <- NA
    if (grepl("(uneingeschränkt)|(voll) (therapie?fh?ähig|behandlungsfähig)|(Patient(in)? Therapie fähig)",
              a_string))
        ret <- 2
    else if (grepl("[Ee]ingeschränk?te? ([Tt]herapie ?[Ff]ähig[keit]?|behandlungsfähig)|(Cis)|((wahrscheinlich)|(fraglich) Therapie fähig)|(TACE)",
                   a_string))
        ret <- 1
    else if (grepl("nicht (therapiefähig|behandlungsfähig)|([Bb]est supportive care)",
                   a_string))
        ret <- 0
    else if (grepl("(unvollständig)|(verstorben)|(fehlt)|(nicht beurteilbar)|(kein geriatrischer Patient)|(Assessment nicht vollständig)",
                   a_string))
        ret <- NA
    ret
}

my_data$beschluss_f <- unlist(sapply(my_data$beschluss, tidy_beschluss))
my_data$beschluss_f <-
    factor(my_data$beschluss_f, ordered = FALSE, levels = c(0:2),
           labels = c("no_go", "slow_go", "go_go"))
```

## Eigene Einschätzung

Vergleich mit Konferenzbeschluss angestrebt
```{r}
my_data$personal_assessment <- as.character(my_data$personal_assessment)
my_data$personal_assessment_f <-
    unlist(sapply(my_data$personal_assessment, tidy_beschluss))
my_data$personal_assessment_f <-
    factor(my_data$personal_assessment_f, ordered = FALSE, levels = c(0:2),
           labels = c("no_go", "slow_go", "go_go"))
summary(my_data$personal_assessment_f)
```

## Geburtsdatum
as.Date Objekt, neue Spalte 'alter' mit Referenz Datum des Konferenzbeschlusses.
```{r}
# neccessary because of internal representation of integers, as.integer(1) is false!
is.wholenumber <- function(x, tol = .Machine$double.eps^0.5)
    ifelse(is.numeric(x), abs(x - round(x)) < tol, FALSE)

geb_datum <- as.integer(as.character(my_data$geburt))

#conversion:
geb_datum <- as.Date(geb_datum, origin = "1899-12-30")
#strange Excel date origin
geb_datum[is.na(geb_datum)] <- as.Date(my_data$geburt[is.na(geb_datum)])
#dates read as strings from Excel, filling the NA's mentioned earlier

my_data$geburt <- geb_datum
rm(geb_datum)
konf_isna <- is.na(my_data$konferenz)
my_data$alter[!konf_isna] <-
    my_data$konferenz[!konf_isna] - my_data$geburt[!konf_isna]

#Wenn kein Konferenzdatum, Aufnahmedatum stattdessen
my_data$alter[konf_isna] <- my_data$aufnahme[konf_isna] - my_data$geburt[konf_isna]

for (f in my_data[is.na(my_data$alter),1]){
    if ((length(grep("2015", f))) > 0)
        my_data$alter[my_data$file_name == f] <-
            as.Date("31.3.2015", format = "%d.%m.%Y") -
            my_data$geburt[my_data$file_name == f]
    else
        if ((length(grep("2014", f))) > 0)
            my_data$alter[my_data$file_name == f] <-
                as.Date("30.6.2015", format = "%d.%m.%Y") -
                my_data$geburt[my_data$file_name == f]
        
}
# alter in days
# conversion to years:
my_data$alter <- trunc(my_data$alter / 365.25)
```

`r sum(is.na(my_data$geburt))` NA's.

```{r}
summary(trunc(as.numeric(my_data$alter / 365)))
hist(as.numeric(my_data$alter))
```

Age is normally distributed.

## IADL

```{r}
my_data$iadl <- as.integer(my_data$iadl)
summary(my_data$iadl)
```

## MMST

```{r}
my_data$mmst <- substr(my_data$mmst, 1, 2) # first two characters
my_data$mmst <- as.integer(my_data$mmst)
summary(my_data$mmst)
```

## DEMTECT

```{r}
my_data$DemTect <- as.numeric(str_replace_all(my_data$DemTect, "[:alpha:]", ""))
summary(my_data$DemTect)
```

## GDS

```{r}
my_data$gds <- as.numeric(as.character(my_data$gds))
summary(my_data$gds)
```

## DIA-S

```{r}
my_data$dia <- as.numeric(as.character(my_data$dia))
summary(my_data$dia)
```

## Timed up and go

If the TUG test is not possible it is a special information, other than if it is only not available. So I define a new column tug_categorical which is factorial, " not possible" is the worst outcome of the test, the number of seconds in the variable tug and NA in both if it is simple not available.

```{r}
my_data$tug_possible[1:length(my_data$tug)] <-
    logical()
my_data$tug_possible[substr(my_data$tug, 1, 5) == "nicht"] <- FALSE
my_data$tug_possible[!substr(my_data$tug, 1, 5) == "nicht"] <- TRUE
my_data$tug <- as.integer(str_extract(my_data$tug, "[0-9]+"))
summary(my_data$tug_possible)
summary(my_data$tug)
```

## Tinetti Test

Format in the original data is 12+5/17, which means first part of the test 12, second part 5, combined score of 17. I have to extract both numbers on either side of the plus sign.

```{r}
my_data$tinetti_1 <- as.integer(str_extract(my_data$tinetti, "[0-9]+"))
# tinetti_1 0..16
my_data$tinetti_2 <- str_extract(my_data$tinetti, "\\++[0-9]+")
my_data$tinetti_2 <- as.integer(substr(my_data$tinetti_2, 2, 4))
# tinetti_2 0..12
# if tinetti is 28 than _1 is 16 an _2 is 12
my_data$tinetti_1[my_data$tinetti=="28"] <- 16
my_data$tinetti_2[my_data$tinetti=="28"] <- 12
my_data$tinetti_1[my_data$tinetti=="100" | my_data$tinetti=="0"] <- NA
my_data$tinetti_2[my_data$tinetti=="100" | my_data$tinetti=="0"] <- NA
summary(my_data$tinetti_1)
summary(my_data$tinetti_2)
my_data$tinetti <- my_data$tinetti_1 + my_data$tinetti_2
```

## Height, Weight, BMI

Mostly formatting and converting.

```{r}
my_data$height <- str_replace_all(my_data$height, "[^0-9.]+", "")
my_data$height <- as.numeric(my_data$height)
my_data$height[!is.na(my_data$height) & my_data$height < 3] <-
    my_data$height[!is.na(my_data$height) & my_data$height < 3] * 100
my_data$height <- as.integer(my_data$height)
summary(my_data$height)
hist(my_data$height)
```

```{r}
my_data$weight <- str_replace_all(my_data$weight, "[^0-9,\\..]+", "")
my_data$weight <- str_replace_all(my_data$weight, ",", ".")
my_data$weight <- str_replace_all(my_data$weight, "\\.{2,2}", ".")
my_data$weight <- round(as.numeric(my_data$weight), 1)
summary(my_data$weight)
hist(my_data$weight)
```

Height and weight are normally distributed!

```{r}
my_data$bmi <- round(my_data$weight/(my_data$height/100)^2, 2)
summary(my_data$bmi)
hist(my_data$bmi)
```

## MNA

```{r}
my_data$mna <- gsub("[^0-9,\\.]", "", my_data$mna)
my_data$mna <- as.numeric(gsub(",", ".", my_data$mna))
summary(my_data$mna)
```

## Charlson Score

```{r}
my_data$charlson <- as.numeric(gsub("[^0-9]", "", my_data$charlson))
summary(my_data$charlson)
```

## Sozialfragebogen

```{r}
my_data$sozial <- as.character(my_data$sozial)

is_living_alone <- function(a_string) {
    ret <- FALSE
    ifelse(grep("(L?lebt allein)|(alleine? lebend)|(Ehemann verstorben)",
                a_string) == 1, ret <- TRUE)
    ret
}

my_data$living_alone <- unlist(lapply(my_data$sozial, is_living_alone))
summary(my_data$living_alone)
```

## Diagnose

```{r}

# my_data$diagnose <- as.character(my_data$diagnose) #nicht mehr noetig, da vorher
# schon mit stringAsFactor = FALSE eingelesen
```

## Indikatorvariablen ob einzelne Tests verfügbar waren
Nötig um später zu entscheiden, ob ein Wert fehlt, weil der Test nicht durchgeführt wurde, oder der Test nicht verfügbar war.
Persönliche Einschätzung ab 1.4.2014
GDS bis 31.8.2014
DIA-S seit 1.9.2014
DemTect seit 1.2.2015
```{r}
my_data$personal_assessment_a <-
    my_data$konferenz>=as.Date("1.4.2014", format="%d.%m.%Y")
my_data$gds_a <-
    my_data$konferenz<=as.Date("31.8.2014", format="%d.%m.%Y")
my_data$dia_a <-
    my_data$konferenz>=as.Date("1.9.2014", format="%d.%m.%Y")
my_data$DemTect_a <-
    my_data$konferenz>=as.Date("1.2.2015", format="%d.%m.%Y")
```

## Conclusion

In the end there is a tidy data frame with one row per patient and one column per category.
```{r}
my_tidy_data <- my_data[,c(-4:-1,-15,-20,-22,-24)]
head(my_tidy_data)

saveRDS(my_tidy_data, "tidy_data.rds")
```
